---
title: "p8106_hw2_sc5826"
author: "Shivalika Chavan"
date: "2026-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(earth)
library(pdp)
```

# Data Cleaning & Splitting into Train/Test
```{r}
college = read_csv("College.csv") |> 
  janitor::clean_names()

set.seed(1234)
indexTrain = createDataPartition(y = college$outstate, p = 0.8, list = FALSE)
college_train = college[indexTrain, ] 
college_test = college[-indexTrain, ] 
```

# A - Smoothing Spline

```{r}
set.seed(1234)
top10perc_grid = seq(min(college_train$top10perc), max(college_train$top10perc), by = 0.25)

spline_df_fits = 
  tibble(df_val = seq(from = 2, to = 52, by = 5)) |> 
  mutate(
    spline_model = map(df_val, \(df) smooth.spline(x = college_train$top10perc, y = college_train$outstate, df = df)),
    predictions = map(spline_model, \(model) predict(model, x = top10perc_grid) |> as.data.frame())
  ) |> 
  unnest(predictions) |> 
  mutate(df_val = as.factor(df_val))

ggplot(college_train, aes(x = top10perc, y = outstate)) +
  geom_point(alpha = 0.5) +
  geom_line(data = spline_df_fits, aes(x = x, y = y, color = df_val)) +
  theme_bw()
```

As the degrees of freedom increase, the smoothing spline line deviates more from a straight line. At the lowest degrees of freedom (`df = 2`), the model has a heavier penalty for roughness, resulting in a nearly straight line. At the other end (`df = 52`), there is a lighter penalty for roughness, and the spline adjusts more to individual points. At higher degrees of freedom, there is a risk of overfitting the training data. 

To select the optimal degrees of freedom, I used the default smoothing spline function, without specifying the degrees of freedome. This calculates and minimizes the default Generalized Cross-Validation (GCV) error to select the tuning parameter ($\lambda$).

```{r}
best_smoothing_spline = smooth.spline(x = college_train$top10perc, y = college_train$outstate)

optimal_df = best_smoothing_spline$df

optimal_pred_df = predict(best_smoothing_spline, x = top10perc_grid) |> 
  as.data.frame()

ggplot(college_train, aes(x = top10perc, y = outstate)) +
  geom_point(alpha = 0.5) +
  geom_line(data = optimal_pred_df, aes(x = x, y = y), color = rgb(.8, .1, .1, 1), linewidth = 1) +
  theme_bw() 
```

Based on the above criterion, the optimal degrees of freedom chosen for the smoothing spline model was ~`r round(optimal_df, 2)`.

# B - MARS

Train a multivariate adaptive regression spline (MARS) model to predict the response
variable. Report the regression function. Present the partial dependence plot of an arbitrary
predictor in your model. Report the test error.

```{r}
# matrix of predictors 
x = model.matrix(outstate ~ . - college, college_train)[, -1]
# vector of response
y = college_train$outstate

mars_grid = expand.grid(degree = 1:3, nprune = 2:30)

set.seed(1234)

ctrl1 = trainControl(method = "cv", 
                     number = 10)

mars_model = train(x, y,
                   method = "earth",
                   tuneGrid = mars_grid,
                   trControl = ctrl1)

ggplot(mars_model)

mars_model$bestTune
```

Regression Function:
```{r}
summary(mars_model$finalModel)
```

Final Regression Function: 

$$
\begin{aligned}
\widehat{Outstate} &= 9221.1968 \\
&+ 0.4164*\text{h(apps-2306)} \\
&- 2.1852*\text{h(1379-f.undergrad)} \\ 
&- 0.4373*\text{h(f.undergrad-1379)} \\ 
&- 1.9481*\text{h(4250-room.board)} \\ 
&+ 51.9742*\text{h(perc.alumni-14)} \\ 
&+ 0.7316*\text{h(expend-5391)} \\ 
&- 0.8081*\text{h(expend-15494)} \\
\end{aligned}
$$

Partial Dependence Plot: 
```{r}
pdp_plot_1 = pdp::partial(mars_model, pred.var = c("f_undergrad"), 
                          grid.resolution = 10) |> 
  autoplot()

pdp_plot_2 = pdp::partial(mars_model, pred.var = c("f_undergrad", "expend"), 
                          grid.resolution = 10) |>
  pdp::plotPartial(levelplot = FALSE, zlab = "yhat", drape = TRUE, screen = list(z = 20, x = -60))

gridExtra::grid.arrange(pdp_plot_1, pdp_plot_2, ncol = 2)
```

Test Error: 
```{r}
predictions = predict(mars_model, newdata = college_test)

test_rmse = RMSE(predictions, college_test$outstate)
```

The test error (RMSE) for the MARS model is `r round(test_rmse, 2)`. 